# Importing data - Part 1
***
Notes taken during/inspired by the Datacamp course 'Importing Data in R (Part 1)' by Filip Schouwenaars.

## Introduction 

Data often comes from many different sources and formats, including

*Flat files - simple tables e.g. csv
*Excel
*Databases - MySQL, Postgres
*Websites - APIs, JSON, scraping
*Other statistical software - SPSS, STATA, SAS

## Reading CSV files

Reading csv files can be achived with simple code like

```{r eval=FALSE}
read.csv("file.csv", stringsAsFactors = FALSE)
```

We may want to import strings as categorical variables, in which case we would set stringsAsFactors = TRUE which is also the default option, if not stated.

When working across different machines or operating systems, problems can arise due to different ways of addressing file locations and differing file locations.  Therefore, it can be easier to set a relative path to the users home directory, which would be achieved with the following code.

```{r}
path <- file.path("~", "datasets", "file.csv")
path
```

Then use the file path as before, assigning to a dataframe.

```{r eval=FALSE}
df <- read.csv(path, stringsAsFactors = FALSE)
```
