# Data Visualization
***
Notes taken during/inspired by the Datacamp course 'Data Visualization in R' by Ronald Pearson.

## Base R Graphics

Two basic types of data visualizations:

* Exploratory visualizations help us understand the data
* Explanatory visualizations help us share ourunderstanding with others 

There are four graphics systems in R:

* Base graphics: Easiest to learn and focus of this course
* Grid graphics: powerful set of modules for building othertools
* Lattice graphics: general purpose system based on grid graphics
* ggplot2: “the grammar of graphics”

Grid graphics allows more control of the output but is more complex, typically it is used by package developers and both lattice and ggplot2 build on this.

```{r}
# Load MASS package
library(MASS)

# Plot whiteside data
plot(whiteside)

# Plot Gas vs. Temp
plot(whiteside$Temp, whiteside$Gas, ylab = "Heating gas consumption", xlab = "Outside temperature")

# Apply the plot() function to Insul - R will automatically change the plot type to suit the data (a factor)
plot(whiteside$Insul)
```

Adding additional details to your explanatory plots can help emphasize certain aspects of your data. For example, by specifying the pch and col arguments to the plot() function, you can add different point shapes and colors to show how different variables or subsets of your data relate to each other. In addition, you can add a new set of points to your existing scatterplot with the points() function, and add reference lines with the abline() function.

```{r}
# Plot Max.Price vs. Price as red triangles
plot(Cars93$Price, Cars93$Max.Price, pch = 17, col = "red")

# Add Min.Price vs. Price as blue circles
points(Cars93$Price, Cars93$Min.Price, pch = 16, col = "blue")

# Add an equality reference line with abline() - lty=2 makes a dashed line
abline(a = 0, b = 1, lty = 2)
```

You can plot multiple graphs on a single pane using the par() function with its mfrow parameter. For example, par(mfrow = c(1, 2)) creates a plot array with 1 row and 2 columns, allowing you to view two graphs side-by-side. This way, you can compare and contrast different datasets or different views of the same dataset. This exercise asks you to compare two views of the Animals2 dataset from the robustbase package, differing in how its variables are represented.

The objective of this exercise is to emphasize that the original representation of the variables that we have in a dataset is not always the best one for visualization or analysis. By representing the original variables in log scale, for example, we can better see and understand the data.

```{r}
# Load the robustbase package
library(robustbase)

# Set up the side-by-side plot array
par(mfrow = c(1, 2))

# First plot: brain vs. body in its original form
plot(Animals2$body, Animals2$brain)

# Add the first title
title("Original representation")

# Second plot: log-log plot of brain vs. body
plot(Animals2$body, Animals2$brain, log = "xy")

# Add the second title
title("Log-log plot")
```

Sometimes, log-transforming the x and y variables by specifying log = "xy" inside the plot() function can make it a lot easier to see the direct relationship between the two variables. Similarly, you can specify log = "y" to log-transform only the y variable and log = "x" to log-transform only the x variable.

### Avoiding pie charts

The same dataset can be displayed or summarized in many different ways, but some are much more suitable than others.

Despite their general popularity, pie charts are often a poor choice. Though R allows pie charts with the pie() function, even the help file for this function argues against their use. Specifically, the help file includes a "Note" that begins with the words: "Pie charts are a very bad way of displaying information."

Bar charts are a recommended alternative.

```{r}
# Load the insuranceData package
library(insuranceData)

# Use the data() function to get the dataCar data frame
data(dataCar)

# Set up a side-by-side plot array
par(mfrow = c(1, 2))

# Create a table of veh_body record counts and sort
tbl <- sort(table(dataCar$veh_body),
            decreasing = TRUE)

# Create the pie chart and give it a title
pie(tbl) + title("Pie chart")

# Create the barplot with perpendicular, half-sized labels
barplot(tbl, las = 2, cex.names = 0.5)

# Add a title
title("Bar chart")
```

## Different Plot Types

Histograms are probably the best-known way of looking at how the values of a numerical variable are distributed over their range, and R provides several different histogram implementations.

The purpose of this exercise is to introduce two of these:

hist() is part of base R and its default option yields a histogram based on the number of times a record falls into each of the bins on which the histogram is based.
truehist() is from the MASS package and scales these counts to give an estimate of the probability density.

```{r}
# Set up a side-by-side plot array
par(mfrow = c(1, 2))

# Create a histogram of counts with hist()
hist(Cars93$Horsepower, main = "hist() plot")

# Create a normalized histogram with truehist()
truehist(Cars93$Horsepower, main = "truehist() plot")
```

While they are probably not as well known as the histogram, density estimates may be regarded as smoothed histograms, designed to give a better estimate of the density function for a random variable.

In this exercise, you'll use the ChickWeight dataset, which contains a collection of chicks' weights. You will first select for the chicks that are 16 weeks old. Then, you'll create a histogram using the truehist() function, and add its density plot on top, using the lines() and density() functions with their default options. The density plot of this type of variable is often expected to conform approximately to the bell-shaped curve, otherwise known as the Gaussian distribution. Let's find out whether that's the case for this dataset.

```{r}
# Create index16, pointing to 16-week chicks
index16 <- which(ChickWeight$Time == 16)

# Get the 16-week chick weights
weights <- ChickWeight$weight[index16]

# Plot the normalized histogram
truehist(weights)

# Add the density curve to the histogram
lines(density(weights))
```


A practical limitation of both histograms and density estimates is that, if we want to know whether the Gaussian distribution assumption is reasonable for our data, it is difficult to tell.

The quantile-quantile plot, or QQ-plot, is a useful alternative: we sort our data, plot it against a specially-designed x-axis based on our reference distribution (e.g., the Gaussian "bell curve"), and look to see whether the points lie approximately on a straight line. In R, several QQ-plot implementations are available, but the most convenient one is the qqPlot() function in the car package.

The first part of this exercise applies this function to the 16-week chick weight data considered in the last exercise, to show that the Gaussian distribution appears to be reasonable here. The second part of the exercise applies this function to another variable where the Gaussian distribution is obviously a poor fit, but the results also show the presence of repeated values (flat stretches in the plot) and portions of the data range where there are no observations (vertical "jumps" in the plot).
```{r}
# Load the car package to make qqPlot() available
library(car)

# Show the normal QQ-plot of the chick weights
qqPlot(weights)

# Show the normal QQ-plot of the Boston$tax data
qqPlot(Boston$tax)
```

The points lie in a relatively straight line only if the data is normally distributed (ie. Its density plot is approximately a Gaussian distribution). So looking at the second QQ-plot, you can see that the tax variable in the Boston data frame is definitely not normally distributed.

A scatterplot represents each (x, y) pair in a dataset by a single point. If some of these pairs are repeated (i.e. if the same combination of x and y values appears more than once and thus lie on top of each other), we can't see this in a scatterplot. Several approaches have been developed to deal with this problem, including jittering, which adds small random values to each x and y value, so repeated points will appear as clusters of nearby points.

A useful alternative that is equally effective in representing repeated data points is the sunflowerplot, which represents each repeated point by a "sunflower," with one "petal" for each repetition of a data point.

Note the large cluster in the upper left corner of the plot, indicating that records with the maximum rad value and zero zn value occur many times.

```{r}
# Set up a side-by-side plot array
par(mfrow = c(1, 2))

# Create the standard scatterplot
plot(rad ~ zn, data = Boston)

# Add the title
title("Standard scatterplot")

# Create the sunflowerplot
sunflowerplot(rad ~ zn, data = Boston)

# Add the title
title("Sunflower plot")
```

The boxplot() function shows how the distribution of a numerical variable y differs across the unique levels of a second variable, x. To be effective, this second variable should not have too many unique levels (e.g., 10 or fewer is good; many more than this makes the plot difficult to interpret).

The boxplot() function also has a number of optional parameters and this exercise asks you to use three of them to obtain a more informative plot:

* varwidth allows for variable-width boxplots that show the different sizes of the data subsets.
* log allows for log-transformed y-values.
* las allows for more readable axis labels.

This exercise also illustrates the use of the formula interface: y ~ x indicates that we want a boxplot of the y variable across the different levels of the x variable.

You can see from the wider boxplots for rad = 4, 5, and 24 how much larger these subsets are than the others.

```{r}
# Create a variable-width boxplot with log y-axis & horizontal labels
boxplot(crim ~ rad, data = Boston, varwidth = TRUE, log = "y", las = 1 )

# Add a title
title("Crime rate vs. radial highway index")
```

A mosaic plot may be viewed as a scatterplot between categorical variables and it is supported in R with the mosaicplot() function.

As this example shows, in addition to categorical variables, this plot can also be useful in understanding the relationship between numerical variables, either integer- or real-valued, that take only a few distinct values.

More specifically, this exercise asks you to construct a mosaic plot showing the relationship between the numerical carb and cyl variables from the mtcars data frame, variables that exhibit 6 and 3 unique values, respectively.

```{r}
# Create a mosaic plot using the formula interface
 mosaicplot(carb ~ cyl, data = mtcars)
```

