# Importing data - Part 1
***
Notes taken during/inspired by the Datacamp course 'Importing Data in R (Part 1)' by Filip Schouwenaars.

## Introduction 

Data often comes from many different sources and formats, including

* Flat files - simple tables e.g. csv
* Excel
* Databases - MySQL, Postgres
* Websites - APIs, JSON, scraping
* Other statistical software - SPSS, STATA, SAS

## Reading CSV files

Reading csv files can be achived with simple code like

```{r eval=FALSE}
read.csv("file.csv", stringsAsFactors = FALSE)
```

We may want to import strings as categorical variables, in which case we would set stringsAsFactors = TRUE which is also the default option, if not stated.

When working across different machines or operating systems, problems can arise due to different ways of addressing file locations and differing file locations.  Therefore, it can be easier to set a relative path to the users home directory, which would be achieved with the following code.

```{r}
path <- file.path("~", "datasets", "file.csv")
path
```

Then use the file path as before, assigning to a dataframe.

```{r eval=FALSE}
df <- read.csv(path, stringsAsFactors = FALSE)
```

## Reading tab deliminated files or other table formats

In a similar way to before, we add the path to the file and if we want strings as strings, for instance

```{r eval=FALSE}
read.delim("file.csv", stringsAsFactors = FALSE)
```

However, if the file comes in another format perhaps due to the system encoding or setup, it is still possible to try and read the file as a tabular formating convering it to a data frame.  To do so, we use the read.table() command which has a lot of arguments that can be customised.  You can specifiy column names and types for instance.  If for instance we have a file format where the objects are seperated by a / rather than a comma or tab as before, we could use

```{r eval = FALSE}
read.table("file.txt",
           header = TRUE,
           sep = "/",
           stringsAsFactors = FALSE)

```

Or, if you have a file which has no column/variable names and tabs as spaces, you would read the file as:

```{r eval = FALSE}
# Path to the file.txt file: path
path <- file.path("data", "file.txt")

# Import the file.txt file: hotdogs
file <- read.table(path, 
                      sep = "\t",                                        # specify seperator - tab in this instance
                      col.names = c("VarName1", "VarName2", "VarName3"), # specifiy variable names
                      colClasses = c("factor", "NULL", "numeric"))       # specify the column/variable classes
```

Both read.csv and read.delim are wrapper functions of read.table(), both use read.table but have different default options depending on the file type.  There are two further wrapper functions - read.csv2 and read.delim2 - which deal with regional differences in formatting, notably that some areas use full stops as decimal places, whereas other areas use commas for decimal places.

## Readr and data.table

These two packages are other ways of reading in files.  Readr uses the tibble, so will be compatable with other tidyverse packages such as dplyr.  It is faster than utils, the r default and also prints out the column classes, depending on what other packages are loaded. It is not neccessary to specifiy stringsAsFactors = FALSE.  

```{r eval = FALSE}
library(readr)
read_csv("file.csv")    #read comma seperated
read_tsv("file2.txt")   #read tab seperated files

#If there are no row heards, you can create a vector then read it in using the col_names argument

#specify the vector for column names
properties <- c("area", "temp", "size", "storage", "method",
                "texture", "flavor", "moistness")
#read in the vector
df <- read_tsv("file3.txt", col_names = properties)

```

Like the utils package, these are wrapper functions, with the base function being read_delim().  Unlike the utils package, read_delim() expects the first row to contain headers, so this doesn't need to be explicit.  As mentioned previously, it is also not neccessary to specify the we don't want strings as factors.  You can specify col_names using a vector as before, or we can read them directly at the time.  If we also want to explicity state the column types, perahps because the automatically assigned variable is not correct, we can do so with col_type using abreviations:

* c = character
* d = double
* i = integer
* n = number
* l = logical
* D = date
* T = date time
* t = time
* ? = guess
* _ = skip column (underscore)

Finally, we can use skip and n_max to specify how many rows to skip at the beggining of a file, perhaps due to a large header, and the maximum now of rows to read, perhaps due to a very large file with many rows. 

```{r eval = FALSE}

read_delim("file4.txt", delim = "/", col_names = c("var1", "var2", "var3"))

read_delim("file5.txt", delim = "/", col_types = "ccid")

read_delim("file6.txt", delim = "\t", col_names = c("var1", "var2", "var3"), 
           skip = 12, n_max = 50000)
```

Another way of setting the types of the imported columns is using collectors. Collector functions can be passed in a list() to the col_types argument of read_ functions to tell them how to interpret values in a column.  Look at the collector documentation for more details.  Two examples are shown below, one for columns to be interpreted as integers and one for a column with factors.

```{r eval = FALSE}
# The collectors needed for importing
fac <- col_factor(levels = c("Beef", "Meat", "Poultry"))
int <- col_integer()

# Edit the col_types argument with the specified collectors
hotdogs_factor <- read_tsv("hotdogs.txt",
                           col_names = c("type", "calories", "sodium"),
                           col_types = list(fac, int, int))

```

### data.table fread

data.table is a tool for doing [fast data analysis](https://github.com/Rdatatable/data.table/wiki/Benchmarks-%3A-Grouping), particularly on large datasets.  It also has a function to read data using the fread() command.  It can autmoatically infer column names, types and seperators.  You can also drop or select columns at read time.

```{r eval = FALSE}
df <- fread("file7.csv", select = c("colname1", "colname2"))
```

The readr package fill create different dataframe types or object classes - 'tbl_df', 'tbl' and 'data.frame' - which can be useful for different purposes, such as for use in dplyr.  Fread creates a data.table object class.