# Intermediate ggplot2
***
Notes taken during/inspired by the Datacamp course 'Data Visualization with ggplot2 (Part 2)' by Rick Scavetta.  This course builds on the first course, which looked at how to build plots, aesthetics, layers and principles.  This second part includes topics such as how to build meaningful plots, statistics and practical matters such as themes.

The focus of this course is onfour layers (statistics, coordinates, facets and themes). 

Course slides:
* [Part 1 - Statistics](https://s3.amazonaws.com/assets.datacamp.com/course/ggplot2/course_2/ggplot2_course2_ch1_slides.pdf)
* [Part 2 - Coordinates and Facets]()
* [Part 3 - Themes]()
* [Part 4 - Best Practices]()
* [Part 5 - Case Study]()

Other useful info:
* [Name](Link)

## Statistics

There are two categories of statistical functions:

* Called from within a geom
* Called independently
* Note geom_FUN <-> stat_FUN

Typically they are called using stat_ but ggplot will translate commands e.g. geom_histogram will call the histogram function. Often warning messages will appear as stat_FUN such as stat_bin when default bin widths are used.  

The following examples use the mtcars data for 32 cars from Motor Trends magazine from 1973. This dataset is small, intuitive, and contains a variety of continuous and categorical (both nominal and ordinal) variables.  Firstly we will add some smoothing lines to the data.  

* You can use either stat_smooth() or geom_smooth() to apply a linear model.

```{r}
library(ggplot2)

# Explore the mtcars data frame with str()
str(mtcars)

# A scatter plot with LOESS smooth:
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth()

# A scatter plot with an ordinary Least Squares linear model:
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm")


# The previous plot, without CI ribbon:
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

# The previous plot, without points:
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_smooth(method = "lm", se = FALSE)
```

We can also look sub-groups in our dataset. For this we'll encounter the invisible group aesthetic.  

* In the first plot, our smooth is calculated for each subgroup because there is an invisible aesthetic, group which inherits from col.  
* In the second plot,  we use multiple aesthetic layers, just like we can use multiple geom layers. Each aesthetic layer can be mapped onto a specific geom.

```{r}
# Define cyl as a factor variable
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = F)

# Complete the following ggplot command as instructed
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = F) + 
  stat_smooth(aes(group = 1), method = "lm", se = F)
```

Here we'll consider the span for LOESS smoothing and we'll take a look at a nice scenario of how to properly map our different models.  LOESS smoothing is a non-parametric form of regression that uses a weighted, sliding-window, average to calculate a line of best fit. We can control the size of this window with the span argument.


Plot 1:  We can control the size of the loess window with the span argument. 

Plot 2: In this plot, we set a linear model for the entire dataset as well as each subgroup, defined by cyl. In the second stat_smooth(),
Set method to "loess" Add span, set it to 0.7

Plot 3: Plot 2 presents a problem because there is a black line on our plot that is not included in the legend. To get this, we need to map something to col as an aesthetic, not just set col as an attribute.
Add col to the aes() function in the second stat_smooth(), set it to "All". This will name the line properly.
Remove the col attribute in the second stat_smooth(). Otherwise, it will overwrite the col aesthetic.

Plot 4: Now we should see our "All" model in the legend, but it's not black anymore.
Add a scale layer: scale_color_manual() with the first argument set to "Cylinders" and values set to the predfined myColors variable

```{r}
# Plot 1: change the LOESS span
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  # Add span below
  geom_smooth(se = F, span = 0.7)

# Plot 2: Set the overall model to LOESS and use a span of 0.7
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = F) +
  # Change method and add span below
  stat_smooth(method = "loess", aes(group = 1),
              se = F, col = "black", span = 0.7)

# Plot 3: Set col to "All", inside the aes layer of stat_smooth()
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = F) +
  stat_smooth(method = "loess",
              # Add col inside aes()
              aes(group = 1, col = "All"),
              # Remove the col argument below
              se = F, span = 0.7)

# Plot 4: Add scale_color_manual to change the colors
myColors <- c(brewer.pal(3, "Dark2"), "black")

ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = F, span = 0.75) +
  stat_smooth(method = "loess",
              aes(group = 1, col="All"),
              se = F, span = 0.7) +
  # Add correct arguments to scale_color_manual
  scale_color_manual("Cylinders", values = myColors)
```

