# Introduction to Data
***
Notes taken during/inspired by the Datacamp course 'Introduction to Data' by Mine Cetinkaya-Rundel.  The supporting textbook is @OS3.

## Language of Data

The course makes use of the openintro package, accompanying the textbook.  Let's load the package and our first dataset, email50.

```{r, message = FALSE}
# Load packages
library("openintro")
library("dplyr")

# Load data
data(email50)
# View its structure
str(email50)
#glimpse the first few items using dplyr
glimpse(email50)

```

When using certain functions, such as filters on categorical variables, the way R handles the filtered out variables is to leave the items in as place holders (empty containers), even though the place holder is empty.  This can have undesirable effects, particularly if using the filtered object for modelling.  We then end up with zero values which are actually filtered out factors.
```{r}
# Subset of emails with big numbers: email50_big
email50_big <- email50 %>%
  filter(number == "big")

# Glimpse the subset
glimpse(email50_big)

# Table of number variable - now we have just 7 values
table(email50_big$number)

# Drop levels
email50_big$number <- droplevels(email50_big$number)

# Another table of number variable
table(email50_big$number)
```

In some instance you want to create a discreet function from a numeric value.  That is to say we want to create a categorical value based on some groups of numbers.  This can be achived as shown below. Note that when calculating a function R will typically either:

* Assign a value e.g. med_num_char <- median(email50$num_char)
* Print a result e.g. median(email50$num_char)
* But we can do both by adding brackets (med_num_char <- median(email50$num_char))

```{r}
# Calculate median number of characters: med_num_char
(med_num_char <- median(email50$num_char))

# Create num_char_cat variable in email50
email50 <- email50 %>%
  mutate(num_char_cat = ifelse(num_char < med_num_char, "below median", "at or above median"))
  
# Count emails in each category
table(email50$num_char_cat)
```

We can also use the mutate function from dplyr to create a new variable from categorical variables

```{r, eval = FALSE}
# Create number_yn column in email50
email50 <- email50 %>%
  mutate(number_yn, ifelse(number == "none", "no", "yes"))

# Visualize number_yn
ggplot(email50, aes(x = number_yn)) +
  geom_bar()

```

We often want to compare two or three variables, which is most easily done using the ggplot package

```{r}
# Load ggplot2
library(ggplot2)

# Scatterplot of exclaim_mess vs. num_char
ggplot(email50, aes(x = num_char, y = exclaim_mess, color = factor(spam))) +
  geom_point()
```

## Observational Studies and Experiments

Typically there are two types of study, if we are interested in whether variable Y is caused by some factors (X) we could have two types of studies.
 
* **Observational Study**: We are observing, rather than specifically interfere or direct how the data is collected - only correlation can be inferred.  In this case, we might survey people and look for patterns in their characteristics (X) and the outcome variable (Y)
* **Experimental Study**: We randomly assign subjects to various treatments - causation can be inferred. In this case, we would get a group of individuals together then randomly assign them to a group of interest (X), removing the decision from the subjects of the study, we often have a control group also.

Another differentiation to be aware of is between 

* **Random sampling**: We select our subjects at random in order that we can make inferences from our sample, to the wider population
* **Random assignment**: Subjects are randomly assigned to various treatments and helps us to make causal conclusions

We can therefore combine random sampling with random assignment, to allow causal and generalisable conclusions, however in practice we typically have one or the other - random sampling only (not causal but generalisable), or random assignment (causal but not generalisable) - the negation of both leads to results that are neither causal nor generalisable, but may highlight a need for further research.

Sometimes when there are looking for associations between variables, it is possible to omit variables of interest, which may be confounding variables.  For instance, we may have two variables (x) that appear to show a relationship with another (y) but the inclusion of a third variable (x') causes the apparent relationship to breakdown.  If we fail to consider other associated variables, we may fall in to a **Simpsons Paradox** in which a trend appears in different groups, but disappears when the groups are combined together.  Simpsons paradox is a form of **Ecological Fallacy**.  One of the best known examples of Simpsons Paradox comes from admissions data for University of California, Berkeley.

```{r}
library(tidyr)
data("UCBAdmissions")
ucb_admit <- as.data.frame(UCBAdmissions)
# Restrucutre data - this is to follow the example provided, it takes the aggregated data from the original data frame and disaggregates 
# it using indexing by repeating the row indices Freq times for each row - see https://stackoverflow.com/questions/45445919/convert-wide-to-long-with-frequency-column
ucb_admit_disagg = ucb_admit[rep(1:nrow(ucb_admit), ucb_admit$Freq), 
                             -grep("Freq", names(ucb_admit))]


# Count number of male and female applicants admitted
ucb_counts <- ucb_admit_disagg %>%
  count(Gender, Admit)

# View result
ucb_counts
  
# Spread the output across columns and calculate percentages
ucb_counts %>%
  spread(Admit, n) %>%
  mutate(Perc_Admit = Admitted / (Admitted + Rejected))

```

So far, it seems that the results suggest females are less likely to be admitted, but what if we look at the results by department?

```{r}
# Table of counts of admission status and gender for each department
admit_by_dept <- ucb_admit_disagg %>%
  count(Dept, Gender, Admit) %>%
  spread(Admit, n)

# View result
admit_by_dept

# Percentage of those admitted to each department
admit_by_dept %>%
  mutate(Perc_Admit = Admitted / (Admitted + Rejected))
```

Now we begin to see that for some departments, there is a higher proportion of females being accpeted than males.  Equally for some departments, the rejection rate is very high for both males and females e.g. Dept F. In 4 of the 6 departments, females have a higher proportion of applications being admitted than males.  Males tended to apply to less competitive departments than females, the less competitive departments had higher admission rates.

## Sampling strategies and experimental design

`r if (knitr:::is_html_output()) '# References {-}'`
