Text Mining in R
***
Notes taken during/inspired by the Datacamp course 'Text Mining: Bag of Words' by Ted Kwartler. 

Course slides:
* [Part 1 - Jumping into text mining with bag of words]()
* [Part 2 - Word clouds and more interesting visuals]()
* [Part 3 - Addining to your skills]()
* [Part 4 - Case Study]()
* [Additional - Dendextend package](https://cran.r-project.org/web/packages/dendextend/vignettes/introduction.html)

## Text Mining Intro

Ted mkaes the definition of Text Mining as 'The process of distilling actionable insights from text'.  We can use Text Mining to reduce the information space in to something more manageable and draw out the important features.  The steps we take are:

* 1 - Problem definition and goals
* 2 - Identify text to be collected
* 3 - Text organisation
* 4 - Feature extraction
* 5 - Aanlysis
* 6 - Reach and insight or make a reccomendation

In this course we look at the bag of words approach, we don't care about word type or order.  The alternative is semantic parsing, which does and looks at word type breaks the elements in to nouns and verb elements, but results in many features for analysis.  

Sometimes we can find out the author's intent and main ideas just by looking at the most common words.

At its heart, bag of words text mining represents a way to count terms, or n-grams, across a collection of documents. 

Manually counting words in the sentences is a pain! Fortunately, the qdap package offers a better alternative. You can easily find the top 4 most frequent terms (including ties) in text by calling the freq_terms function and specifying 4.

> frequent_terms <- freq_terms(text, 4)

The frequent_terms object stores all unique words and their count. You can then make a bar chart simply by calling the plot function on the frequent_terms object.

> plot(frequent_terms)

```{r}
new_text <- "DataCamp is the first online learning platform that focuses on building the best learning experience specifically for Data Science. We have offices in Boston and Belgium and to date, we trained over 250,000 (aspiring) data scientists in over 150 countries. These data science enthusiasts completed more than 9 million exercises. You can take free beginner courses, or subscribe for $25/month to get access to all premium courses."

# Load qdap
library(qdap)

# Print new_text to the console
new_text

# Find the 10 most frequent terms: term_count
term_count <- freq_terms(new_text, 10)

# Plot term_count
plot(term_count)
```

Next we are going to build our first corpus of text using tweets.  A corpus is a collection of documents, specifying it as a text source for the tm package.  

Text mining begins with loading some text data into R, which we'll do with the read.csv() function. By default, read.csv() treats character strings as factor levels like Male/Female. To prevent this from happening, it's very important to use the argument stringsAsFactors = FALSE.

Best practice is to examine the object you read in to make sure you know which column(s) are important. The str() function provides an efficient way of doing this. You can also count the number of documents using the nrow() function on the new object. In this example, it will tell you how many coffee tweets are in the vector.

If the data frame contains columns that are not text, you may want to make a new object using only the correct column of text (e.g. some_object$column_name).

```{r}
# Import text data
tweets <-  read.csv("D:/CloudStation/Documents/2017/RData/coffee.csv", stringsAsFactors = FALSE) 

# View the structure of tweets
str(tweets)

# Print out the number of rows in tweets
nrow(tweets)

# Isolate text from tweets: coffee_tweets
coffee_tweets <- tweets$text
```

Your next step is to convert this vector containing the text data to a corpus, a corpus is a collection of documents, but it's also important to know that in the tm domain, R recognizes it as a data type.

There are two kinds of the corpus data type, the permanent corpus, PCorpus, and the volatile corpus, VCorpus. In essence, the difference between the two has to do with how the collection of documents is stored in your computer. In this course, we will use the volatile corpus, which is held in your computer's RAM rather than saved to disk, just to be more memory efficient.

To make a volatile corpus, R needs to interpret each element in our vector of text, coffee_tweets, as a document. And the tm package provides what are called Source functions to do just that! In this exercise, we'll use a Source function called VectorSource() because our text data is contained in a vector. The output of this function is called a Source object.

```{r}
# Load tm
library(tm)

# Make a vector source: coffee_source
coffee_source <- VectorSource(coffee_tweets)
```

Now that we've converted our vector to a Source object, we pass it to another tm function, VCorpus(), to create our volatile corpus. 

The VCorpus object is a nested list, or list of lists. At each index of the VCorpus object, there is a PlainTextDocument object, which is essentially a list that contains the actual text data (content), as well as some corresponding metadata (meta). It can help to visualize a VCorpus object to conceptualize the whole thing.

For example, to examine the contents of the second tweet in coffee_corpus, you'd subset twice. Once to specify the second PlainTextDocument corresponding to the second tweet and again to extract the first (or content) element of that PlainTextDocument:

> coffee_corpus[[15]][1]

```{r}
# Make a volatile corpus: coffee_corpus
coffee_corpus <- VCorpus(coffee_source)

# Print out coffee_corpus
coffee_corpus

# Print data on the 15th tweet in coffee_corpus
coffee_corpus[[15]]

# Print the content of the 15th tweet in coffee_corpus
coffee_corpus[[15]]$content
```

Because another common text source is a data frame, there is a Source function called DataframeSource(). The DataframeSource() function treats the entire row as a complete document, so be careful you don't pick up non-text data like customer IDs when sourcing a document this way.

```{r}
# Setup the example text
num <- 1:3
Author1 <- c("Text mining is a great time.","Text analysis provides insights" , "qdap and tm are used in text mining")
Author2 <- c("R is a great language" , "R has many uses" , "DataCamp is cool!")
example_text <- cbind(num, Author1, Author2)
example_text <- as.data.frame(example_text, stringsAsFactors = FALSE)

# Print example_text to the console
example_text

# Create a DataframeSource on columns 2 and 3: df_source
df_source <- DataframeSource(example_text[, 2:3])

# Convert df_source to a corpus: df_corpus
df_corpus <- VCorpus(df_source)

# Examine df_corpus
df_corpus

# Create a VectorSource on column 3: vec_source
vec_source <- VectorSource(example_text$Author2)

# Convert vec_source to a corpus: vec_corpus
vec_corpus <- VCorpus(vec_source)

# Examine vec_corpus
vec_corpus

```

### Cleaning and Preprocessing Text

There are a number of functions within R that can help process text, including tolower(), removePunctuation(), removeNumbers(), stripWhiteSpace() and removeWords() - however it is important to note and consider that some of these elements might convey meaning, so removing them can strip some of that meaning out.  

In bag of words text mining, cleaning helps aggregate terms. For example, it may make sense that the words "miner", "mining" and "mine" should be considered one term. Specific preprocessing steps will vary based on the project. For example, the words used in tweets are vastly different than those used in legal documents, so the cleaning process can also be quite different.

Common preprocessing functions include:

* tolower(): Make all characters lowercase
* removePunctuation(): Remove all punctuation marks
* removeNumbers(): Remove numbers
* stripWhitespace(): Remove excess whitespace

Note that tolower() is part of base R, while the other three functions come from the tm package. Going forward, we'll load the tm and qdap for you when they are needed. Every time we introduce a new package, we'll have you load it the first time.

```{r}
# Create the object: text
text <- "<b>She</b> woke up at       6 A.M. It\'s so early!  She was only 10% awake and began drinking coffee in front of her computer."

# All lowercase
tolower(text)

# Remove punctuation
removePunctuation(text)

# Remove numbers
removeNumbers(text)

# Remove whitespace
stripWhitespace(text)
```

The qdap package offers other text cleaning functions. Each is useful in its own way and is particularly powerful when combined with the others.

* bracketX(): Remove all text within brackets (e.g. "It's (so) cool" becomes "It's cool")
* replace_number(): Replace numbers with their word equivalents (e.g. "2" becomes "two")
* replace_abbreviation(): Replace abbreviations with their full text equivalents (e.g. "Sr" becomes "Senior")
* replace_contraction(): Convert contractions back to their base words (e.g. "shouldn't" becomes "should not")
* replace_symbol() Replace common symbols with their word equivalents (e.g. "$" becomes "dollar")

```{r}
## text is still loaded in your workspace

# Remove text within brackets
bracketX(text)

# Replace numbers with words
replace_number(text)

# Replace abbreviations
replace_abbreviation(text)

# Replace contractions
replace_contraction(text)

# Replace symbols with words
replace_symbol(text)
```

Often there are words that are frequent but provide little information. So you may want to remove these so-called stop words. Some common English stop words include "I", "she'll", "the", etc. In the tm package, there are 174 stop words on this common list.

In fact, when you are doing an analysis you will likely need to add to this list. In our coffee tweet example, all tweets contain "coffee", so it's important to pull out that word in addition to the common stop words. Leaving it in doesn't add any insight and will cause it to be overemphasized in a frequency analysis.

Using the c() function allows you to add new words (separated by commas) to the stop words list. For example, the following would add "word1" and "word2" to the default list of English stop words:

> all_stops <- c("word1", "word2", stopwords("en"))

Once you have a list of stop words that makes sense, you will use the removeWords() function on your text. removeWords() takes two arguments: the text object to which it's being applied and the list of words to remove.

```{r}
# List standard English stop words
stopwords("en")

# Print text without standard stop words
removeWords(text, stopwords('en'))

# Add "coffee" and "bean" to the list: new_stops
new_stops <- c("coffee", "bean", stopwords("en"))

# Remove stop words from text
removeWords(text, new_stops)
```

Still another useful preprocessing step involves word stemming and stem completion. The tm package provides the stemDocument() function to get to a word's root. This function either takes in a character vector and returns a character vector, or takes in a PlainTextDocument and returns a PlainTextDocument.

For example,

> stemDocument(c("computational", "computers", "computation"))

returns "comput" "comput" "comput". But because "comput" isn't a real word, we want to re-complete the words so that "computational", "computers", and "computation" all refer to the same word, say "computer", in our ongoing analysis.

We can easily do this with the stemCompletion() function, which takes in a character vector and an argument for the completion dictionary. The completion dictionary can be a character vector or a Corpus object. Either way, the completion dictionary for our example would need to contain the word "computer" for all the words to refer to it.

```{r}
# Create complicate
complicate <- c("complicated", "complication", "complicatedly")

# Perform word stemming: stem_doc
stem_doc <- stemDocument(complicate)

# Create the completion dictionary: comp_dict
comp_dict <- "complicate"

# Perform stem completion: complete_text 
complete_text <- stemCompletion(stem_doc, comp_dict)

# Print complete_text
complete_text
```

Let's consider the following sentence as our document for this exercise:

> "In a complicated haste, Tom rushed to fix a new complication, too complicatedly."

This sentence contains the same three forms of the word "complicate" that we saw in the previous exercise. The difference here is that even if you called stemDocument() on this sentence, it would return the sentence without stemming any words. 

This happens because stemDocument() treats the whole sentence as one word. In other words, our document is a character vector of length 1, instead of length n, where n is the number of words in the document. To solve this problem, we first remove the punctation marks with the removePunctuation(). We then strsplit() this character vector of length 1 to length n, unlist(), then proceed to stem and re-complete.

```{r}
text_data <- "In a complicated haste, Tom rushed to fix a new complication, too complicatedly."

# Remove punctuation: rm_punc
rm_punc <- removePunctuation(text_data)

# Create character vector: n_char_vec
n_char_vec <- unlist(strsplit(rm_punc, split = ' '))

# Perform word stemming: stem_doc
stem_doc <- stemDocument(n_char_vec)

# Print stem_doc
stem_doc

# Re-complete stemmed document: complete_doc
complete_doc <- stemCompletion(stem_doc, comp_dict) 

# Print complete_doc
complete_doc
```

The tm package provides a special function tm_map() to apply cleaning functions to a corpus. Mapping these functions to an entire corpus makes scaling the cleaning steps very easy.

To save time (and lines of code) it's a good idea to use a custom function, since you may be applying the same functions over multiple corpora. Here we use the clean_corpus() function. It takes one argument, corpus, and applies a series of cleaning functions to it in order, then returns the final result.

Notice how the tm package functions do not need content_transformer(), but base R and qdap functions do.

Be sure to test your function's results. If you want to draw out currency amounts, then removeNumbers() shouldn't be used. Plus, the order of cleaning steps makes a difference. For example, if you removeNumbers() and then replace_number(), the second function won't find anything to change.

```{r}
# Alter the function code to match the instructions
clean_corpus <- function(corpus){
  corpus <- tm_map(corpus, stripWhitespace)
  corpus <- tm_map(corpus, removePunctuation)
  corpus <- tm_map(corpus, content_transformer(tolower))
  corpus <- tm_map(corpus, removeWords, c(stopwords("en"), "coffee", "mug"))
  return(corpus)
}

# Apply your customized function to the tweet_corp: clean_corp
clean_corp <- clean_corpus(coffee_corpus)

# Print out a cleaned up tweet
clean_corp[[227]][1]

# Print out the same tweet in original form
tweets$text[227]
```

In bag of words analysis, either we have a Term Document Matrix (TDM) or a Document Term Matrix (DTM).  One is a transposition of the other, 

* TDM - rows are terms/words, columns are tweets/documents
* DTM - rows are tweets/documents, columns are terms/words

Let's revisit the coffee tweets to build a document-term matrix.

Beginning with the coffee.csv file, we have used common transformations to produce a clean corpus called clean_corp.

The document-term matrix is used when you want to have each document represented as a row. This can be useful if you are comparing authors within rows, or the data is arranged chronologically and you want to preserve the time series.

```{r}
# Create the dtm from the corpus: coffee_dtm
coffee_dtm <- DocumentTermMatrix(clean_corp)

# Print out coffee_dtm data
coffee_dtm

# Convert coffee_dtm to a matrix: coffee_m
coffee_m <- as.matrix(coffee_dtm) 

# Print the dimensions of coffee_m
dim(coffee_m)

# Review a portion of the matrix
coffee_m[148:150, 2587:2590]
```

Next we are performing a similar process but taking the transpose of the document-term matrix. In this case, the term-document matrix has terms in the first column and documents across the top as individual column names.

The TDM is often the matrix used for language analysis. This is because you likely have more terms than authors or documents and life is generally easier when you have more rows than columns. An easy way to start analyzing the information is to change the matrix into a simple matrix using as.matrix() on the TDM.

```{r}
# Create a TDM from clean_corp: coffee_tdm
coffee_tdm <- TermDocumentMatrix(clean_corp)

# Print coffee_tdm data
coffee_tdm

# Convert coffee_tdm to a matrix: coffee_m
coffee_m <- as.matrix(coffee_tdm)

# Print the dimensions of the matrix
dim(coffee_m)

# Review a portion of the matrix
coffee_m[2587:2590, 148:150]
```

## Word Clouds and other visuals

Visual information is quickly processed so using visuals helps.  Here we are going to look at two corpora - one for coffee tweets and the other for chardonnay tweets.  

Calling rowSums() on your newly made matrix aggregates all the terms used in a passage. Once you have the rowSums(), you can sort() them with decreasing = TRUE, so you can focus on the most common terms.

Lastly, you can make a barplot() of the top 5 terms of term_frequency with the following code.

> barplot(term_frequency[1:5], col = "#C0DE25")

```{r}
# Create a matrix: coffee_m
coffee_m <- as.matrix(coffee_tdm)

# Calculate the rowSums: term_frequency
term_frequency <- rowSums(coffee_m) 

# Sort term_frequency in descending order
term_frequency <- sort(term_frequency,
 decreasing = TRUE) 

# View the top 10 most common words
term_frequency[1:10]

# Plot a barchart of the 10 most common words
barplot(term_frequency[1:10],
 col = "tan", las = 2)
```

A fast way to get frequent terms is with freq_terms() from qdap, although you loose some control.  The function accepts a text variable, which in our case is the tweets$text vector. You can specify the top number of terms to show with the top argument, a vector of stop words to remove with the stopwords argument, and the minimum character length of a word to be included with the at.least argument. qdap has its own list of stop words that differ from those in tm. Our exercise will show you how to use either and compare their results.

Making a basic plot of the results is easy. Just call plot() on the freq_terms() object.

Note that using different lists of stop words can produce different results, as shown here.

```{r}
# Create frequency
frequency <- freq_terms(
 tweets$text,
 top = 10,
 at.least = 3,
 stopwords = "Top200Words"
 ) 

# Make a frequency barchart
plot(frequency)

# Create frequency2
frequency2 <- freq_terms(
 tweets$text,
 top = 10,
 at.least = 3,
 stopwords = tm::stopwords("english")
 ) 

# Make a frequency2 barchart
plot(frequency2)
```

### Word Clouds

n celebration of making it this far, let's try our hand on another batch of 1000 tweets. For now, you won't know what they have in common, but let's see if you can figure it out using a word cloud. The tweets' term-document matrix, matrix, and frequency values are preloaded in your workspace.

A word cloud is a visualization of terms. In a word cloud, size is often scaled to frequency and in some cases the colors may indicate another measurement. For now, we're keeping it simple: size is related to individual word frequency and we are just selecting a single color.

As you saw in the video, the wordcloud() function works like this:

> wordcloud(words, frequencies, max.words = 500, colors = "blue")

Text mining analyses often include simple word clouds. In fact, they are probably over used, but can still be useful for quickly understanding a body of text.

```{r}
# Load the data
tweets2 <-  read.csv("D:/CloudStation/Documents/2017/RData/chardonnay.csv", stringsAsFactors = FALSE) 

# Isolate text from tweets: chard_tweets
chard_tweets <- tweets2$text

# Make a vector source: coffee_source
chard_source <- VectorSource(chard_tweets)

# Make a volatile corpus: coffee_corpus
chard_corpus <- VCorpus(chard_source)

# Alter the function code to match the instructions
clean_corpus <- function(corpus){
  corpus <- tm_map(corpus, stripWhitespace)
  corpus <- tm_map(corpus, removePunctuation)
  corpus <- tm_map(corpus, content_transformer(tolower))
  corpus <- tm_map(corpus, removeWords, c(stopwords("en"),  "amp",
 "chardonnay", "wine", "glass"))
  return(corpus)
}

# Apply your customized function to the tweet_corp: clean_chardonnay
clean_chardonnay <- clean_corpus(chard_corpus)

# Create a TDM from clean_corp: chardonnay_tdm
chardonnay_tdm <- TermDocumentMatrix(clean_chardonnay)

# Create a matrix: chardonnay_m
chardonnay_m <- as.matrix(chardonnay_tdm)

# Sum rows and sort by frequency
term_frequency <- rowSums(chardonnay_m)
word_freqs <- data.frame(term = names(term_frequency),
 num = term_frequency)

# Load wordcloud package
library(wordcloud)

# Print the first 10 entries in term_frequency
term_frequency[1:10]

# Create word_freqs
word_freqs <- data.frame(
  term = names(term_frequency),
  num = term_frequency
)

# Create a wordcloud for the values in word_freqs
wordcloud(
  word_freqs$term, 
  word_freqs$num,
  max.words = 100, 
  colors = "red")
```

