# Writing Functions in R
***
Notes taken during/inspired by the Datacamp course 'Writing Functions in R' by Hadley and Charlotte Wickham.

R can create functions using a basic 'receipe' as Hadley calls it


```{r, eval = FALSE}
my_fun <- function(arg1, arg2) {
   body
}
```

> 

Unlike in other languages, there is no special syntax for naming a function. Once created, there is no difference between the functions you create and the functions created by other developers such as those in base R.  

Every function has three components

* The formal arguments
* The body of the function 
* The environment - usually invisible but determines where the function looks for variables

The environemnt is where the function was defined, if just coding the function in this will be the global environment.  

The output from a function is usually the last expression evaluated.  However, it is possible to use a return statement to return(value) which will stop the function at that point.  It is possible to have functions without a name, so called 'anonymous' functions, which are covered later in this chapter.  Anonymous functions can be called, but have to be done on the same line.  

Course slides:
* [Part 1 - Refresher](https://s3.amazonaws.com/assets.datacamp.com/production/course_1008/slides/ch1_pdf_slides_v2.pdf)
* [Part 2 - When and how you should write a function]()
* [Part 3 - Functional programming]()
* [Part 4 - Advanced inputs and outputs]()
* [Part 5 - Robust functions]

## Function overview






## When and how you should write a function

Why might you write a function? Doing the same thing many times is not efficient and it can lead to errors - you copy the same function/formula from a previous exercise, but forget to change a variable name to fit your new instance.  

> If you have copied and pasted twice, meaning you now have three copies, it is time to write a function

As it takes less effort to check the intent of the code - there is just one function to check - more time can be spent on checking the validity or quality of the data and outputs.  In addition, since we have created a function, we can use other packages such as the map function in purrr, to repeat (map) our functions repeatadly.  

### Rescale example

We have a snippet of code that successfully rescales a column to be between 0 and 1:

  (df$a - min(df$a, na.rm = TRUE)) /  
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
  
Our goal over the next few exercises is to turn this snippet, written to work on the a column in the data frame df, into a general purpose rescale01() function that we can apply to any vector.

The first step of turning a snippet into a function is to examine the snippet and decide how many inputs there are, then rewrite the snippet to refer to these inputs using temporary names. These inputs will become the arguments to our function, so choosing good names for them is important. (We'll talk more about naming arguments in a later exercise.)

In this snippet, there is one input: the numeric vector to be rescaled (currently df$a). What would be a good name for this input? It's quite common in R to refer to a vector of data simply as x (like in the mean function), so we will follow that convention here.

```{r}
# Define example vector x
x <- seq(1:10)

# Rewrite this snippet to refer to x
(x - min(x, na.rm = TRUE)) /
  (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
```

Our next step is to examine our snippet and see if we can write it more clearly.

Take a close look at our rewritten snippet. Do you see any duplication?

  (x - min(x, na.rm = TRUE)) /
  (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
  
One obviously duplicated statement is min(x, na.rm = TRUE). It makes more sense for us just to calculate it once, store the result, and then refer to it when needed. In fact, since we also need the maximum value of x, it would be even better to calculate the range once, then refer to the first and second elements when they are needed.

What should we call this intermediate variable? You'll soon get the message that using good names is an important part of writing clear code! I suggest we call it rng (for "range").

```{r}
# Define example vector x
x <- 1:10

# Define rng
rng <- range(x, na.rm = TRUE)

# Rewrite this snippet to refer to the elements of rng
(x - rng[1]) /
  (rng[2] - rng[1])
```

What do you need to write a function? You need a name for the function, you need to know the arguments to the function, and you need code that forms the body of the function.

```{r}
# Define example vector x
x <- 1:10 

# Use the function template to create the rescale01 function
rescale01 <- function(x) {
  # body
  rng <- range(x, na.rm = TRUE) 
  (x - rng[1]) / (rng[2] - rng[1])
}

# Test your function, call rescale01 using the vector x as the argument
rescale01(x)
```

